Parsing mPFC: A Joint Meta-Analytic and Graph Theoretic Approach

### 2018-09-21 Manuscript.Rmd
Mostly advancing on results (day and session analyses)
But some changes to the script so the code from these sections matches better.

In order to evaluate the stability and topographical heterogeneity of the partitions within and across subjects, we chose the adjusted rand index (ARI) [@Hubert1985] as the reference metric of agreement between two clusterings. The adjusted rand index estimates the rate at which every pair of nodes in partition A were equally assigned to the same/different parcels on partition B, over all possible changes (e.g. clustered together on A, but separately on B). The resulting ratio is compared to a baseline given by the expectation based on a random vertex assignment for an equal number of clusters across partitions. Because of this, the adjusted rand index is a simple metric that determines the percentage of agreement between any two partitions that is agnostic to the labeling scheme. Note that this metric is adjusted for chance cluster labelings, and thus heavily penalizes deviations.

@@

With these 16 ROIs (15 overlapping, 1 DMN-specific), we estimated each individual's DMN coverage using the whole time series (4800 total TRs from four scanning sessions over two days). Each subject's network (n = `r nSubj`) was constructed by performing pairwise correlations among the time series of all the surface vertices contained within the ROIs (`r nrow(Summaries[[1]])` total vertices). After correcting for multiple testing, we performed spectral partitioning and modularity on these networks to identify cohesive functional communities denoting DMN and non-DMN areas. On average, networks had high densities (i.e. highly connected) regardless of the amount of data to produce them (mean = `r round(mdAll, digits = 2)`, SD = `r sddAll`). This degree of density is important, as these methods are known to face challenges when computed on sparse networks [@Fortunato2016].

Figure 2 shows a stereotypical subdivision of the ROIs for a single subject (100307) using spectral partitioning. This method dissects a network by the positive and negative values derived from the Fiedler vector (the eigenvector related to the second-to-lowest eigenvalue of the network's Laplacian matrix, see Methods), which we used to binarize the partition. Since this is a data-driven approach, there is no a priori labeling for these communities. We thus defined the DMN community as the one containing the one DMN-specific area from the meta-analysis (7m), and ensured that the binary labeling of the vertices was consistent across subjects. In general, the resulting pattern shows large DMN coverage in PCC, with non-DMN vertices populating its anterior segment. mPFC, on the other hand, is mostly populated by DMN vertices in its ventral-anterior and dorsal-anterior areas, with a persistent non-DMN pocket between them. This non-DMN section extends posteriorly onto pregenual cinglate cortex. Note that vmPFC does not encompass area a25, which is often related to negative emotional valuation [@Joyce2018]. Interstingly, this topographical combination of dorsal and ventral mPFC, with extensive PCC coverage, matches previous meta-analytic patterns more closely attributed to social cognition than SV [@Hiser2018].

@@

Before determining the degree of generazibility of this topological pattern across individuals, we examined the validity of these partitions by comparing them to patterns derived with a modularity algorithm [@Clauset2004]. Modularity seeks to find the pattern of communities that maximizes a quality function that compares the connectivity strength of any two nodes with a null probability model. Since modularity is agnostic to the desired number of clusters, it was capable of finding more than two communities in our datset. Thus, if both methods agreed in the number and topological organization of the current communities, it would add confidence to our results. We quantified the cross-method agreement with the adjusted rand index (ARI). ARI measures the proportion of pairs of nodes in a network that were either clustered together or separately in both partitions, over all possible affiliation changes (e.g. clustered together in partition 1, but separately in 2), and has the benefit of being agnostic to labeling schemes [@Hubert1985]. Figure 3 shows the distribution of ARI values across subjects.

@@

HCP resting state data is separated into 4 runs (2 per day of scanning). This section explores how much agreement there is between partitions computed on each day. The similarity matrix on the left shows the pairwise level of agreement for the overall data among the example subjects, while the right one divides this data into PCC and mPFC.

@@

So far both visualizations show considerable subject specificity, and the regional division agrees with the whole-data similarity matrix. Importantly, this indicates that both PCC and mPFC partitions were more similar within- than between-subjects (e.g. RI values higher for each 2x2 diagonal submatrix than the off-diagonals). Now, how do we check whether the within subject consistency is significantly higher than across subjects? One way is to divide each subject's RI-agreement between days by the RI values with the days from the other subjects. This gives an RI ratio per subject that denotes the relative specificity of their partition. That means, partitions that align well across subjects should yield ratios close to 1, while subject-consistent and unique partitions would produce high ratios.

``` {r Partition comparison: halves ratios (Fig. 6), fig.align="center", fig.width=3, fig.height=3, fig.cap = "Figure 6. mPFC displays higher within-subject agreement than between subjeccts, as evidenced by a high within/between subject mean ARI. PCC also shows some level of topographical idiosyncrasy (ratio > 1), but significantly lower than mPFC.", echo=FALSE, message = FALSE, cache = TRUE}
# Now compare all possible pairs of subjects
# Compare all halves with RI
RI_all_between_halves <- comparePartitions(Data = Summaries_halves, Index = "RI", nSubjects = length(Summaries_halves), subjNames = subjList_halves)

@@

## Ratio of mean within subject sessions over between subject ones (per-subject)
# Selection of submatrices
template <- diagBlocks(ncol(RI_all_between_halves), 2) 
lowTemplate <- template[lower.tri(template)]

# Overall
# Split the within-subject session RI values per subject
withinSubjVals <- RI_all_between_halves[lower.tri(RI_all_between_halves)][lowTemplate]

@@

# PCC
withinSubjVals <- RI_PCC_halves[lower.tri(RI_PCC_halves)][lowTemplate]

@@

# PFC
withinSubjVals <- RI_PFC_halves[lower.tri(RI_PFC_halves)][lowTemplate]

@@

# Stats
Halves_mPFCvPCC_permTest <- permute(Halves_RIRatios_PFC, Halves_RIRatios_PCC)
Halves_mPFCvPCC_ES <- cohenD(Halves_RIRatios_PFC, Halves_RIRatios_PCC)

# plot
ggplot(data = Halves_RIRatios_summary, aes(Region, RI, fill = Region)) + 
  geom_violin(show.legend = F, draw_quantiles = 0.5, trim = F, lwd = 0.7) +
  scale_fill_manual(values = Cols[seq(2)]) + 
  ylim(0, NA) +
  geom_hline(yintercept = 1) +
  labs(y = "ARI Ratio", x = "") + 
  theme_classic(base_size = 14)
```

As expected, we see that while both regions had good specificity (ratios > 1, denoted by the horizontal line), the ratios for PCC were significantly smaller than mPFC for all subjects (permutation, p = `r Halves_mPFCvPCC_permTest$Pval`; Cohen's D = `r Halves_mPFCvPCC_ES`). However, the denominator is given by a single RI. If these topological patterns are indeed subject-specific for mPFC, these results might generalize to session specific data. This would give us a better estimate of within-subject RI variance to place in the denominator.

@@

Next, each session is evaluated independently. The similarity matrix on the left shows how each session's data corresponded within and between subjects for all ROIs combined, while the one on the right shows PCC and mPFC separately. It is worth noting that each session is ~14 mins long, which is below the stability threshold for fMRI based modularity (Gordon et al., 2017). However, we also perform a sliding window analysis below that encompasses 20 mins of data per window.

@@

As before, we see higher agreement levels within subjects (with some exceptions). Note that the overall ARI has markedly diminished for mPFC, which is expected as the 15 minute sessions are too short to determine stable community patterns [@Gordon2017]. Nevertheless, we can still coarsely capture a subject's unique organizational pattern. Simple permutation comparisons confirmed this intuition for the all-ROI (p = `r Sess_overallComparison[[3]]$Pval`; Cohen's D `r round(Sess_overallComparison[[4]], digits = 2)`), PCC (p = `r Sess_PCCComparison[[3]]$Pval`; Cohen's D = `r round(Sess_PCCComparison[[4]], digits = 2)`), and mPFC (p = `r Sess_PFCComparison[[3]]$Pval`; Cohen's D = `r round(Sess_PFCComparison[[4]], digits = 2)`) divisions. In order to quantify the consistency and specificity of these partitions, we again applied the RI ratio index, but this time using the mean RI among all of a subject's sessions in the denominator. The following plot shows the results for PCC and mPFC separately.

@@

# Stats
Sess_mPFCvPCC_permTest <- permute(Sess_RIRatios_PFC, Sess_RIRatios_PCC)
Sess_mPFCvPCC_ES <- cohenD(Sess_RIRatios_PFC, Sess_RIRatios_PCC)

@@

We can see again that the ratios for mPFC were higher than those of PCC (permutation, p = `r Sess_mPFCvPCC_permTest$Pval`; Cohen's D = `r round(Sess_mPFCvPCC_ES, digits = 2)`), but both were above 1. This suggests that the PCC is well aligned across subjects (as mentioned before), while mPFC is consistent within subject, but quite different from other subjects.

Another way to determine the heterogeneity of these results is based on Kragel et al. (2018), who quantified the dissimilarity of brain patterns for behavioral domain, subdomain, and study (or how generalizeable the patterns are at each hierarchical level). In our case, for `r nSubj` subjects `r nSubj*3` matrices are created (`r nSubj*2` for daily comparisons, `r nSubj` for within subject comparisons). Each matrix is binarized to compare a subject's level of similarity within their daily data vs across every other subject's sessions. The `r nSubj*3` matrices are vectorized and used as predictors in a GLM with a quasibinomial distribution, where the dependent data is the full vectorized RI similarity matrix. This is a simplified version of a fully-dimensional model that considers all possible comparisons at the session level, which would otherwise have S + (S x (S-1)/2) parameters (with S = session). For this model, a significantly negative subject predictor implies that the odds of the partitions being similar across subjects are much lower than those within each subject (across sessions). Thus, subject specificity would dictate that all subject-level coefficients should be significant and negative.

@@

It is also worth considering how stable these measures are, so we performed a sliding window analysis  per subject (20 min windows shifting at ~1 min throughout the whole ~ 1hr time series). The first two plots show how well each window's clustering agreed with the partition derived from the whole data (in RIs, left), as well as how variable (in standard deviations) this index was throughout each time series (right).

@@

# Mean RI across windows per subject
SC_summary <- aggregate(SC_all$Index, by = list(SC_all$SubjID), FUN = mean)
SC_summary$SD <- aggregate(SC_all$Index, by = list(SC_all$SubjID), FUN = sd)$x
colnames(SC_summary) <- list("SubjID", "RI", "SD")
SC_sd_plot <- ggplot(data = SC_summary, aes(x=SD)) + 
  geom_density(color = "black", fill = Cols[2], bins=50) + 
  xlim(0,1) +
  labs(x = "SD") +
  theme(text = element_text(size=35)) +
  theme_classic()

>>>>

In order to evaluate the stability and topographical heterogeneity of the partitions within and across subjects, we chose the adjusted rand index (ARI) as the reference metric of agreement between two clusterings [@Hubert1985]. ARI estimates the rate at which every pair of nodes in partition A were jointly assigned to the same/different parcels on partition B, over all possible changes (e.g. clustered together on A, but separately on B). Because of this, the adjusted rand index determines the percentage of agreement between any two partitions while being agnostic to the labeling scheme. The resulting ratio is compared to a baseline given by the expectation of a random node assignment across partitions. This means that even though clustering deviations are heavily penalized, positive ARI values compare favorably against chance clustering (and can take negative values if the index falls below chance).

@@

With these 16 ROIs (15 overlapping, 1 DMN-specific), we estimated each individual's DMN coverage using the whole time series (4800 total TRs from four scanning sessions over two days). Each subject's network (n = `r nSubj`) was constructed by performing pairwise correlations among the time series of all the surface vertices contained within the ROIs (`r nrow(Summaries[[1]])` total vertices). After correcting for multiple testing, we performed spectral partitioning and modularity on these networks to identify cohesive functional communities denoting DMN and non-DMN areas. On average, networks had high densities (i.e. highly connected) regardless of the amount of data to produce them (mean = `r round(mdAll, digits = 2)`, SD = `r round(sddAll, digits = 3)`). This degree of density is important, as these methods are known to face challenges when computed on sparse networks [@Fortunato2016].

Figure 2 shows a stereotypical subdivision of the ROIs for a single subject (100307) using spectral partitioning. This method dissects a network by the positive and negative values derived from the Fiedler vector (the eigenvector related to the second-to-lowest eigenvalue of the network's Laplacian matrix, see Methods), which we used to binarize the ROIs Since this is a data-driven approach, there is no a priori labeling for these communities. We thus defined the DMN community as the one containing the one DMN-specific area from the meta-analysis (7m), and ensured that the binary labeling of the vertices was consistent across subjects. In general, the resulting pattern shows large DMN coverage in PCC, with non-DMN vertices populating its anterior segment. mPFC, on the other hand, is mostly covered by DMN vertices in its ventral-anterior and dorsal-anterior areas, with a persistent non-DMN pocket between them. This non-DMN section extends posteriorly onto pregenual cinglate cortex. Note that vmPFC does not encompass area a25, which is often related to negative emotional valuation [@Joyce2018]. Interstingly, this topographical combination of dorsal and ventral mPFC, with extensive PCC coverage, matches previous meta-analytic patterns more closely attributed to social cognition than SV [@Hiser2018].

@@

Before determining the degree of generazibility of this topological pattern across individuals, we examined the validity of these partitions by comparing them to patterns derived with a modularity algorithm [@Clauset2004]. Modularity seeks to find the pattern of communities that maximizes a quality function that compares the connectivity strength of any two nodes with a null probability model. Since modularity is agnostic to the desired number of clusters, it was capable of finding more than two communities in our datset. Thus, if both methods agreed in the number and topological organization of the current communities, it would add confidence to our results. We quantified the cross-method agreement with the adjusted rand index (ARI). ARI measures the proportion of pairs of nodes in a network that were either clustered together or separately in both partitions, over all possible affiliation changes (e.g. clustered together in partition 1, but separately in 2). This measure has the benefit of being agnostic to labeling schemes, as well as controlling for the expectation that the clusters were randomly produced [@Hubert1985]. This means that positive ARI values compare favorably against chance clustering. Figure 3 shows the distribution of ARI values comparing topologies produced by modularity and spectral partitioning across subjects.

@@

In order to determine whether the low agreement shown in mPFC across subjects was a function of intrinsic variability or a subject-specific pattern, we compared the agreement among communities estimated with data from each separate day. If the latter were true, within-subject ARI values should be consistently higher than across subjects. Since SP and modularity showed converging estimates, from here on we focus on results computed with spectral partitioning to ensure bipartitions. Communities were estimated on concatenated pairs of daily left-right and right-left phase encoded sessions (see Methods for details), and once again compared using ARI. Figure 5 shows pairwise comparisons among ten example subjects for PCC and mPFC separately.

@@

``` {r Partition comparison: halves ratios (Fig. 6), fig.align="center", fig.width=3, fig.height=3, fig.cap = "Figure 6. mPFC displays higher within-subject agreement than between subjeccts, as evidenced by a high within/between subject mean ARI. PCC also shows some level of topographical idiosyncrasy (ratio > 1), but significantly lower than mPFC.", echo=FALSE, message = FALSE, cache = TRUE}
## Now compare all possible pairs of subjects
# Compare all halves with RI
RI_all_between_halves <- comparePartitions(Data = Summaries_halves, Index = "RI", nSubjects = length(Summaries_halves), subjNames = subjList_halves)

@@

## Let's compare stuff (not accounting for complex interdependencies in the observations)
## Overall differences in mean RI between PCC and mPFC
lowPCC_halves <- RI_PCC_halves[lower.tri(RI_PCC_halves)]
lowPFC_halves <- RI_PFC_halves[lower.tri(RI_PFC_halves)]
Halves_PCCvPFC_perm <- permute(lowPCC_halves, lowPFC_halves, statType = mean, paired = T)
Halves_PCCvPFC_ES <- cohenD(lowPCC_halves, lowPFC_halves) # caveat: maybe not ideal for bounded values, like RI. BUT VI yields similar results.

## Differences in RI values for sessions within vs between subjects (all subjects pooled together)
# create a selection matrix containing all within-subject session comparisons
template <- diagBlocks(ncol(RI_all_between_halves), 2) 
lowTemplate <- template[lower.tri(template)]

# Overall
# First list element is the within subject values, second is the between subject values
lowRIHalves <- RI_all_between_halves[lower.tri(RI_all_between_halves)]
Halves_overallComparison <- list(lowRIHalves[lowTemplate],
                                lowRIHalves[!lowTemplate])
Halves_overallComparison[[3]] <- permute(Halves_overallComparison[[1]], Halves_overallComparison[[2]])
Halves_overallComparison[[4]] <- cohenD(Halves_overallComparison[[1]], Halves_overallComparison[[2]])

# PCC 
Halves_PCCComparison <- list(lowPCC_halves[lowTemplate],
                           lowPCC_halves[!lowTemplate])
Halves_PCCComparison[[3]] <- permute(Halves_PCCComparison[[1]], Halves_PCCComparison[[2]])
Halves_PCCComparison[[4]] <- cohenD(Halves_PCCComparison[[1]], Halves_PCCComparison[[2]])

# mPFC
Halves_PFCComparison <- list(lowPFC_halves[lowTemplate],
                           lowPFC_halves[!lowTemplate])
Halves_PFCComparison[[3]] <- permute(Halves_PFCComparison[[1]], Halves_PFCComparison[[2]])
Halves_PFCComparison[[4]] <- cohenD(Halves_PFCComparison[[1]], Halves_PFCComparison[[2]])

## Ratio of mean within subject sessions over between subject ones (per-subject)
# Overall
# Split the within-subject session RI values per subject
withinSubjVals <- RI_all_between_halves[lower.tri(RI_all_between_halves)][lowTemplate]

@@

# PCC
# Split the within-subject session RI values per subject
withinSubjVals <- RI_PCC_halves[lower.tri(RI_PCC_halves)][lowTemplate]

@@

# PFC
# Split the within-subject session RI values per subject
withinSubjVals <- RI_PFC_halves[lower.tri(RI_PFC_halves)][lowTemplate]

@@

# Stats
Halves_mPFCvPCC_permTest <- permute(Halves_RIRatios_PFC, Halves_RIRatios_PCC, paired = T)
Halves_mPFCvPCC_ES <- cohenD(Halves_RIRatios_PFC, Halves_RIRatios_PCC)

# plot
ggplot(data = Halves_RIRatios_summary, aes(Region, RI, fill = Region)) + 
  geom_violin(show.legend = F, draw_quantiles = 0.5, trim = F, lwd = 0.7) +
  scale_fill_manual(values = Cols[seq(2)]) + 
  ylim(0, 10) +
  geom_hline(yintercept = 1) +
  labs(y = "ARI Ratio", x = "") + 
  theme_classic(base_size = 14)

Once again we found greater alignment across subjects for PCC (mean = `r round(mean(Halves_PCCComparison[[2]]), digits = 2)`, SD = `r round(sd(Halves_PCCComparison[[2]]), digits = 2)`) than mPFC (mean = `r round(mean(Halves_PFCComparison[[2]]), digits = 2)`, SD = `r round(sd(Halves_PFCComparison[[2]]), digits = 2)`), but both areas showed higher levels of agreement within subjects (PCC: mean = `r round(mean(Halves_PCCComparison[[1]]), digits = 2)`, SD = `r round(sd(Halves_PCCComparison[[1]]), digits = 2)`; mPFC: mean = `r round(mean(Halves_PFCComparison[[1]]), digits = 2)`, SD = `r round(sd(Halves_PFCComparison[[1]]), digits = 2)`). However, a coarse comparison of mean ARIs neglects the degree to which subject-wise partitions were indeed idiosyncratic. Therefore, we divided each subject's ARI between days by the ARI values with the days from the other subjects. This ratio denotes the relative specificity of their partition. That means, partitions that align well across subjects should yield ratios close to 1, while subject-consistent and unique partitions would produce high ratios. Figure 6 shows the ratios for PCC and mPFC for all subjects. As expected, we found that while both regions had good specificity (i.e. ratios > 1), the ratios for PCC were significantly lower than mPFC (paired permutation, p = `r Halves_mPFCvPCC_permTest$Pval`; Cohen's D = `r Halves_mPFCvPCC_ES`). These test/retest results suggest that the topographic varability seen in mPFC is rooted in subject-specific organization, and not an inability to capture its topology.

@@

We extended the analysis of daily data by examining whether the organization of the DMN could be extracted using session data only. It is worth noting that the length of each session (approx. 14 mins long) falls below the stability threshold for fMRI-based modularity estimations [@Gordon2017]. Nonetheless, high ARI ratios with reduced data could signify the ability to localize a subject's DMN organization with fMRI sessions that are easier to fit along with other study sessions of interest.

@@

Computing SP on networks from individual sessions shows that we can still coarsely capture a subject's unique organizational pattern (Figure 7), even though the overall levels of agreement decreased (PCC between subjects: mean = `r round(mean(Sess_PCCComparison[[2]]), digits = 2)`, SD = `r round(sd(Sess_PCCComparison[[2]]), digits = 2)`; PCC within subjects: mean = `r round(mean(Sess_PCCComparison[[1]]), digits = 2)`, SD = `r round(sd(Sess_PCCComparison[[1]]), digits = 2)`); mPFC between subjects: mean = `r round(mean(Sess_PFCComparison[[2]]), digits = 2)`, SD = `r round(sd(Sess_PFCComparison[[2]]), digits = 2)`; mPFC within subjects: mean = `r round(mean(Sess_PFCComparison[[1]]), digits = 2)`, SD = `r round(sd(Sess_PFCComparison[[1]]), digits = 2)`). We again computed each subject's ARI ratio in order to quantify the specificity of these partitions, but this time using the mean ARI acros sessions (Figure 8).

@@

# Stats
Sess_mPFCvPCC_permTest <- permute(Sess_RIRatios_PFC, Sess_RIRatios_PCC, paired = T)
Sess_mPFCvPCC_ES <- cohenD(Sess_RIRatios_PFC, Sess_RIRatios_PCC)

@@

We can see again that the ratios for mPFC were higher than those of PCC (permutation, p = `r Sess_mPFCvPCC_permTest$Pval`; Cohen's D = `r round(Sess_mPFCvPCC_ES, digits = 2)`), but both were mostly above 1. This further confirms that the intrinsic functional organization of mPFC is uniquely arranged per individual, and provides evidence that such patterns can be identified with relatively small amounts of data in spite of a reduction in the estimates.

**Another way to determine the heterogeneity of these results is based on Kragel et al. (2018), who quantified the dissimilarity of brain patterns for behavioral domain, subdomain, and study (or how generalizeable the patterns are at each hierarchical level). In our case, for `r nSubj` subjects `r nSubj*3` matrices are created (`r nSubj*2` for daily comparisons, `r nSubj` for within subject comparisons). Each matrix is binarized to compare a subject's level of similarity within their daily data vs across every other subject's sessions. The `r nSubj*3` matrices are vectorized and used as predictors in a GLM with a quasibinomial distribution, where the dependent data is the full vectorized RI similarity matrix. This is a simplified version of a fully-dimensional model that considers all possible comparisons at the session level, which would otherwise have S + (S x (S-1)/2) parameters (with S = session). For this model, a significantly negative subject predictor implies that the odds of the partitions being similar across subjects are much lower than those within each subject (across sessions). Thus, subject specificity would dictate that all subject-level coefficients should be significant and negative. **

@@

While the distinction of DMN and non-DMN is consistent across session and days, it is possible that the decrease in global ARI values going from full to session-specific data could be due to a dynamic shift in organization. In order to address this possibility, we performed a sliding window analysis on each subject's full time series (20 min windows shifting at 1 min), comparing the community arrangement derived on each window with the original pattern estimated using the whole time series. The first two plots show how well each window's clustering agreed with the partition derived from the whole data (in RIs, left), as well as how variable (in standard deviations) this index was throughout each time series (right).

@@

# Mean RI across windows per subject
SC_summary <- aggregate(SC_all$Index, by = list(SC_all$SubjID), FUN = mean)
SC_summary$SD <- aggregate(SC_all$Index, by = list(SC_all$SubjID), FUN = sd)$x
colnames(SC_summary) <- list("SubjID", "RI", "SD")
SC_sd_plot <- ggplot(data = SC_summary, aes(x=SD)) + 
  geom_density(color = "black", fill = Cols[2]) + 
  xlim(0,1) +
  labs(x = "SD") +
  theme(text = element_text(size=35)) +
  theme_classic()
